Endogenous Macrodynamics in Algorithmic Recourse

### 2022-08-09 dev/paper/sections/introduction.qmd
minor things

To the best of our knowledge this is the first work investigating endogenous macrodynamics in AR. Our contributions to the state of knowledge are as follows: firstly, we posit a compelling argument that calls for a novel perspective on algorithmic recourse extending our focus from single individuals to groups. Secondly, we introduce an experimental framework extending previous work by @altmeyer2022CounterfactualExplanations, which enables us to study macrodynamics of algorithmic recourse through simulations. To this end we propose a number of novel evaluation metrics that can be used to qunatify and benchmark the macrodynmics introduced by different counterfactual generators. Thirdly, we use this framework to provide a first in-depth analysis of endogenous recourse dynamics induced by various popular counterfactual generators including @wachter2017counterfactual, @schut2021generating, @joshi2019towards, @mothilal2020explaining and @antoran2020getting. Finally, we discuss what drives endogenous dynamics and strategies to mitigate them.

>>>>

To the best of our knowledge this is the first work investigating endogenous macrodynamics in AR. Our contributions to the state of knowledge are as follows: firstly, we posit a compelling argument that calls for a novel perspective on algorithmic recourse extending our focus from single individuals to groups. Secondly, we introduce an experimental framework extending previous work by @altmeyer2022CounterfactualExplanations, which enables us to study macrodynamics of algorithmic recourse through simulations that can be fully parallelized. Thirdly, we use this framework to provide a first in-depth analysis of endogenous recourse dynamics induced by various popular counterfactual generators including @wachter2017counterfactual, @schut2021generating, @joshi2019towards, @mothilal2020explaining and @antoran2020getting. To this end we propose a number of novel evaluation metrics that can be used to quantify and benchmark the macrodynamics introduced by the different generators. Finally, we also discuss what drives endogenous dynamics and propose strategies to mitigate them.

### 2022-08-09 dev/paper/sections/related.qmd
minor things

@@

Despite the large and growing number of approaches to counterfactual search there have been surprisingly few benchmark studies that compare different methodologies. This may be partially due to limited software availability in this space. Recent work has started to address this gap: firstly, @de2021framework run a large benchmarking study using different algorithmic aproaches and numerous tabular datasets; secondly, @pawelczyk2021carla introduce a Python framework that can be used to apply and benchmark different methodology; finally, @altmeyer2022CounterfactualExplanations provides an extensible and language-agnostic implementation in Julia. For the experiments conducted in this work we relied on the latter.

>>>>

As mentioned earlier, research has so far also generally focused on generating counterfactuals for single individuals or instances. We have been able to identify only one existing work that investigates black-box model behavior towards a group of individuals [@carrizosa2021generating]. The authors propose an optimization framework that generates collective counterfactuals. We provide a motivation for doing so from the perspective of endogenous macrodynamics of algorithmic recourse.

@@

Despite the large and growing number of approaches to counterfactual search there have been surprisingly few benchmark studies that compare different methodologies. This may be partially due to limited software availability in this space. Recent work has started to address this gap: firstly, @de2021framework run a large benchmarking study using different algorithmic aproaches and numerous tabular datasets; secondly, @pawelczyk2021carla introduce a Python framework that can be used to apply and benchmark different methodologies; finally, @altmeyer2022CounterfactualExplanations provides an extensible, fast and language-agnostic implementation in Julia. Since the experiments presented here involve extensive simulations, we have relied on and extended the Julia implementation due to the associated performance benefits.
